Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals

### 2021-11-22 09_Article/01_Revision2_Reviewers/Simon_2021_Bone_Manuscript_Revised.tex
Implement Philippe suggestions into manuscript

		Compared to healthy controls, we found the OI samples to have significantly lower BV/TV and trabecular number (Tb.N.), significantly higher CV, trabecular separation (Tb.Sp.) and trabecular separation standard deviation (Tb.Sp.SD), but no differences in trabecular thickness (Tb.Th.). These results are in agreement with previous studies. The stiffnesses of highly heterogeneous ROIs were randomly lower with respect to the fabric-elasticity relationships, which reflects the limit of validity of the computational homogenisation methodology. \Add{Nevertheless, the error performed on these highly heterogeneous ROIs is negligible in comparison with homogeneous ROIs which are order of magnitude stiffer.}

@@

	\subsubsection{Model Modifications}
	Another modification of the model was to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. A similar mixed-effect model was used to analyze the relation between TMD and BV/TV and the significance of the group, see Appendix \ref{A1} Equation \ref{EqA14}. The model used the BV/TV and the group (healthy or OI) as fixed variables and the individual as random variable. Moreover, to test the hypothesis of no interaction between the BV/TV and the group, i.e. the group has no significant influence on the TMD versus BV/TV slope, the model was modified to add the interaction regressor (BV/TV x Group). The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}.\\

@@

	As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This influences the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. Finally, the significant differences observed in BV/TV and DA, even with matched age and sex, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\

	The linear regressions performed in this study on original data sets showed $R^2_{adj}$ and NE in the expected range (i. e. slightly lower than \citeauthor{Gross2013}\cite{Gross2013} and \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}) for the healthy group. Components of the stiffness tensors are distributed to both sides of the diagonal. On the other hand, the linear regressions performed in this study using the original OI data set presents lower $R^2_{adj}$ and higher NE than comparable linear regressions in literature \cite{Gross2013,Panyasantisuk2015}. The important value of NE and its standard deviation shows that the fitted stiffness can deviate significantly from the observation. These differences come from ROIs presenting a low stiffness. The linear regression plot (Figure \ref{02_OI}) shows that when the stiffness term decreases to \Add{1 MPa (}10\textsuperscript{0}\Add{)} \Del{MPa} and lower, the linear regression tends to overestimate the stiffness. This is because ROI stiffness is highly dependent on BV/TV values. Some ROIs with low BV/TV do not have every side of the cube connected by bone, leading to extremely low terms in the stiffness tensor (see Appendix \ref{A2}). Trying to homogenise such ROIs can lead to errors of multiple order of magnitude, as observed on the plot (Figure \ref{02_OI}). Therefore, a filtering is indispensable to assess and compare fabric-elasticity relationships, as done by \citeauthor{Panyasantisuk2015} \cite{Panyasantisuk2015}. An alternative to CV filtering for assessing the ROI heterogeneity could be to compute the proportion of the area filled by bone on each of the six faces of the ROI. \\

	Figure \ref{02_CV_BVTV} presents the CV in relation to BV/TV. It shows a tendency of CV values to increase with decreasing BV/TV values. Effectively, if the quantity of material inside the ROI decreases, the distribution homogeneity of this mass is more sensitive and therefore can quickly become highly heterogeneous. A simple assumption about this relation is that it could be monotonic. Pearson's correlation coefficient being strictly negative confirms a negative monotonic relation. As some ROIs with higher BV/TV still present high CV values, imposing a fixed threshold for subsequent homogenisation seems feasible. However, the value of this threshold results from an optimisation process in the study of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and could be investigated further. \Add{Nevertheless, it should be noted that this relation between BV/TV and CV as well as the homogenization process (i.e. fabric-elasticity relationships) are independent of the OI type. Distribution of OI types for CV in relation to BV/TV as well as the linear regression of the original data set are available in Appendix} \ref{A3}\Add{.} \\

	The linear regressions performed on the filtered data sets present a direct effect of filtering as the ROIs meeting the homogeneity assumption lead to better results compared to linear regression including ROIs with high CV values. For the healthy group (N=603), the relatively small decrease of $R^2_{adj}$ (0.5\%) compared to the linear regression using the unfiltered data set is negligible. On the other hand, NE values are decreased by 2\% and therefore improved. The filtering eliminates data points further away from the diagonal (better NE) and other data points close to the diagonal leading to a smaller number of points (modifying $R^2_{adj}$). For the OI group, filtering leads to an important improvement of the linear regression (i.e. higher $R^2_{adj}$ and lower NE). $R^2_{adj}$, NE, and the range of stiffness values are almost at the level of the healthy group. These results give confidence to the filtering procedure and represent a first step in accepting the hypothesis of healthy and OI trabecular bone having the same fabric-elasticity relationships.\\

	After BV/TV and DA matching, grouping the data sets together led to similar $R^2_{adj}$ and NE as for the individual filtered data sets. This allows one to determine values for $k$ and $l$ for the tibia at a spatial resolution of 61 \si{\micro}m. Imposing these values to perform the linear regression on data sets of the matched individuals allows us to highlight differences, if any, between healthy and OI trabecular bone. The relatively low differences for $\lambda_0'$ and $\mu_0$ once again supports the hypothesis for similar fabric-elasticity relationships between healthy and OI trabecular bone. For $\lambda_0$, this relative difference being higher could rise some doubts about this similarity, but the 95\% CI intervals still show a common range which almost include both the $\lambda_0$ of OI and healthy linear regressions. Moreover, ANCOVA performed comparing the original formulation and the one with addition of a regressor for the group showed a p-value far above the 5\% significance level. With this statistical non-significance of the groups and their low relative differences in the computed stiffness constants, it can be stated that: if trabecular bone is homogeneous enough, there is no reason to assume differences in fabric-elasticity relationships between healthy and OI trabecular bone. Thus, the same hFE methodology can be applied for bone stiffness assessment of OI diagnosed individuals as for healthy ones\Add{, regardless the OI type}. Although in hFE analysis it is not possible to exclude part of the mesh because of high heterogeneity, the influence of such ROIs seems negligible on full bone models as the fabric-elasticity relationships can extrapolate their stiffness smoothly and as this concerns only ROIs with low BV/TV and stiffness.\Add{ Moreover, ROI side length can be reduced to improve accuracy if necessary for trabecular bone presenting extremely low BV/TV (and thus high heterogeneity).}\\

@@

	The main limitations of this study are the definition of "homogeneous enough" (i.e. low CV) and the fact that it is limited to XCTII reconstructions of the distal tibia. Moreover, having only one patient with OI type III where we could extract ROIs does not allow to perform statistics on OI types. ROI homogeneity, more specifically a balanced bone connectivity of the 6 faces of the ROI, is a necessary condition underlying the computational homogenisation of mechanical properties and the results of heterogenous samples must be disregarded. However, when calibrated with homogeneous ROIs, fabric-elas\-ticity relationships offer a smooth and reasonable extrapolation of stiffness properties for the concerned low-homogeneity and low-stiffness ROIs. In fact, these low-property ROIs were not shown to compromise the performance of hFE models of the distal radius \cite{Varga2011, Hosseini2017, AriasMoreno2019}.\\

@@

	In conclusion, OI trabecular bone from the distal tibia analysed in the present study had similar morphology compared to those reported in the literature. In particular, we confirmed a significantly lower (-26\%) volume fraction of OI trabecular bone compared to healthy controls that points towards a substantial reduction in bone strength of 45\% when applying the well-established squared relationship with BV/\-TV \cite{Fleps2020}. 
	We also observed a 5\% higher TMD in OI % With mean of each group 4.3% with fitted line at mean BV/TV range 4.7% TMD increase in OI.
	which suggests, according to the results of \citeauthor{Indermaur2021}\cite{Indermaur2021}, an increase in matrix ultimate stress of about 13\%. % Mean BV/TV:0.249 healthy: 576, OI: 603, Michi slope: y = -0.4 + 0.0011*x
	However, this moderate increase in matrix properties is large\-ly overcompensated by the 45\% reduction in trabecular bone strength associated with BV/TV. % Using k=2: (74**2-100**2)/100**2
	We did not find statistically significant differences in fabric-elasticity relationships between healthy and OI trabecular bone, when only valid, homogenous ROIs were included and matched for both BV/TV and DA. Therefore, HR-pQCT-based hFE analyses used for bone stiffness estimation of healthy individuals can be used for OI diagnosed patients, potentially with a small correction for TMD.

>>>>

		Compared to healthy controls, we found the OI samples to have significantly lower BV/TV and trabecular number (Tb.N.), significantly higher CV, trabecular separation (Tb.Sp.) and trabecular separation standard deviation (Tb.Sp.SD), but no differences in trabecular thickness (Tb.Th.). These results are in agreement with previous studies. The stiffnesses of highly heterogeneous ROIs were randomly lower with respect to the fabric-elasticity relationships, which reflects the limit of validity of the computational homogenisation methodology. \Add{This limitation does not challenge the fabric-elasticity relationship, which extrapolation to heterogeneous ROIs is probably reasonable but can simply not be evaluated with the employed homogenisation methodology. Moreover, due to their low BV/TV, the potential (unknown) errors on these heterogeneous ROIs would have negligible influence on whole bone stiffness in comparison to homogeneous ROIs which are orders of magnitude stiffer.}

@@

	\subsubsection{Model Modifications}
	Another modification of the model was to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. A similar mixed-effect model was used to analyze the relation between TMD and BV/TV and the significance of the group, see Appendix \ref{A1} Equation \ref{EqA14}. The model used the BV/TV and the group (healthy or OI) as fixed variables and the individual as random variable. Moreover, to test the hypothesis of no interaction between the BV/TV and the group, i.e. the group has no significant influence on the TMD versus BV/TV slope, the model was modified to add the interaction regressor (BV/TV x Group). The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}. \\

@@

	As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This influences the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. \Add{In fact, high heterogeneity is found at the edge of void spaces or medullary cavities that are much larger in OI compared to healthy controls.} Finally, the significant differences observed in BV/TV and DA, even with matched age and sex, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\

	The linear regressions performed in this study on original data sets showed $R^2_{adj}$ and NE in the expected range (i. e. slightly lower than \citeauthor{Gross2013}\cite{Gross2013} and \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}) for the healthy group. Components of the stiffness tensors are distributed to both sides of the diagonal. On the other hand, the linear regressions performed in this study using the original OI data set presents lower $R^2_{adj}$ and higher NE than comparable linear regressions in literature \cite{Gross2013,Panyasantisuk2015}. The important value of NE and its standard deviation shows that the fitted stiffness can deviate significantly from the observation. These differences come from ROIs presenting a low stiffness. The linear regression plot (Figure \ref{02_OI}) shows that when the stiffness term decreases to \Add{1 MPa (}10\textsuperscript{0}\Add{)} \Del{MPa} and lower, the linear regression tends to overestimate the stiffness. This is because ROI stiffness is highly dependent on BV/TV values. 	Some ROIs with low BV/TV do not have every side of the cube connected by bone, leading to extremely low terms in the stiffness tensor (see Appendix \ref{A2}). Trying to homogenise such ROIs can lead to errors of multiple order of magnitude, as observed on the plot (Figure \ref{02_OI}). \Del{Therefore}\Add{This represents a limitation of the homogenisation methodology and} a filtering \Add{according to the heterogeneity level becomes}\Del{is} indispensable to assess and compare fabric-elasticity relationships, as done by \citeauthor{Panyasantisuk2015} \cite{Panyasantisuk2015}. An alternative to CV filtering for assessing the ROI heterogeneity could be to compute the proportion of the area filled by bone on each of the six faces of the ROI. \\

	Figure \ref{02_CV_BVTV} presents the CV in relation to BV/TV. It shows a tendency of CV values to increase with decreasing BV/TV values. Effectively, if the quantity of material inside the ROI decreases, the distribution homogeneity of this mass is more sensitive and therefore can quickly become highly heterogeneous. A simple assumption about this relation is that it could be monotonic. Pearson's correlation coefficient being strictly negative confirms a negative monotonic relation. As some ROIs with higher BV/TV still present high CV values, imposing a fixed threshold for subsequent homogenisation seems feasible. However, the value of this threshold results from an optimisation process in the study of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and could be investigated further. \Add{Nevertheless, it should be noted that this relation between BV/TV and CV as well as the fabric-elasticity relationships are independent of the OI type. Distribution of OI types for CV in relation to BV/TV as well as the linear regression of the original data set are available in Appendix \ref{A3}.} \\

	The linear regressions performed on the filtered data sets present a direct effect of filtering as the ROIs meeting the homogeneity assumption lead to better results compared to linear regression including ROIs with high CV values. For the healthy group (N=603), the relatively small decrease of $R^2_{adj}$ (0.5\%) compared to the linear regression using the unfiltered data set is negligible. On the other hand, NE values are decreased by 2\% and therefore improved. The filtering eliminates data points further away from the diagonal (better NE) and other data points close to the diagonal leading to a smaller number of points (modifying $R^2_{adj}$). For the OI group, filtering leads to an important improvement of the linear regression (i.e. higher $R^2_{adj}$ and lower NE). $R^2_{adj}$, NE, and the range of stiffness values are almost at the level of the healthy group. These results give confidence to the filtering procedure and represent a first step in accepting the hypothesis of healthy and OI trabecular bone having the same fabric-elasticity relationships. \Add{Indeed, this confirms that filtering according to heterogeneity level allows to compare fabric-elasticity relationships of homogeneous ROIs for which the selected homogenisation procedure is valid.}\\

	After BV/TV and DA matching, grouping the data sets together led to similar $R^2_{adj}$ and NE as for the individual filtered data sets. This allows one to determine values for $k$ and $l$ for the tibia at a spatial resolution of 61 \si{\micro}m. Imposing these values to perform the linear regression on data sets of the matched individuals allows us to highlight differences, if any, between healthy and OI trabecular bone. The relatively low differences for $\lambda_0'$ and $\mu_0$ once again supports the hypothesis for similar fabric-elasticity relationships between healthy and OI trabecular bone. For $\lambda_0$, this relative difference being higher could rise some doubts about this similarity, but the 95\% CI intervals still show a common range which almost include both the $\lambda_0$ of OI and healthy linear regressions. Moreover, ANCOVA performed comparing the original formulation and the one with addition of a regressor for the group showed a p-value far above the 5\% significance level. With this statistical non-significance of the groups and their low relative differences in the computed stiffness constants, it can be stated that: if trabecular bone is homogeneous enough, there is no reason to assume differences in fabric-elasticity relationships between healthy and OI trabecular bone. Thus, the same hFE methodology can be applied for bone stiffness assessment of OI diagnosed individuals as for healthy ones\Add{, regardless the OI type}. Although in hFE analysis it is not possible to exclude part of the mesh because of high heterogeneity, the influence of such ROIs seems negligible on full bone models as the fabric-elasticity relationships can extrapolate their stiffness smoothly and as this concerns only ROIs with low BV/TV and stiffness\Del{.}\Add{ orders of magnitude lower than the homogeneous ones or the cortex (< 1 MPa compared to 10 GPa). Moreover, studies validated the use of fabric-elasticity relationships in osteoporotic bone \cite{Varga2011,Hosseini2017} or vertebral bodies with metastatic defects \cite{Stadelmann2020} where high heterogeneity is present.}\\
	%\Add{ Moreover, ROI side length can be reduced to improve accuracy if necessary for trabecular bone presenting extremely low BV/TV (and thus high heterogeneity).}\\

@@

	The main limitations of this study are the definition of "homogeneous enough" (i.e. low CV) and the fact that it is limited to XCTII reconstructions of the distal tibia. Moreover, having only one patient with OI type III where we could extract ROIs does not allow to perform statistics on OI types. \Add{Nevertheless, the lack of homogeneous type III ROIs is essentially due to the lack of trabecular bone in which case no relationship is needed. No outlier and no evidence of a different relationship was seen in our results for any OI type including the sample with OI type III.} ROI homogeneity, more specifically a balanced bone connectivity of the 6 faces of the ROI, is a necessary condition underlying the computational homogenisation of mechanical properties and the results of heterogenous samples must be disregarded. However, when calibrated with homogeneous ROIs, fabric-elas\-ticity relationships offer a smooth and reasonable extrapolation of stiffness properties for the concerned low-homogeneity and low-stiffness ROIs. In fact, these low-property ROIs were not shown to compromise the performance of hFE models of the distal radius \cite{Varga2011, Hosseini2017, AriasMoreno2019}.\\

@@

	In conclusion, OI trabecular bone from the distal tibia analysed in the present study had similar morphology compared to those reported in the literature. In particular, we confirmed a significantly lower (-26\%) volume fraction of OI trabecular bone compared to healthy controls that points towards a substantial reduction in bone strength of 45\% when applying the well-established squared relationship with BV/\-TV \cite{Fleps2020}. 
	We also observed a 5\% higher TMD in OI % With mean of each group 4.3% with fitted line at mean BV/TV range 4.7% TMD increase in OI.
	which suggests, according to the results of \citeauthor{Indermaur2021}\cite{Indermaur2021}, an increase in matrix ultimate stress of about 13\%. % Mean BV/TV:0.249 healthy: 576, OI: 603, Michi slope: y = -0.4 + 0.0011*x
	However, this moderate increase in matrix properties is large\-ly overcompensated by the 45\% reduction in trabecular bone strength associated with BV/TV. % Using k=2: (74**2-100**2)/100**2
	We did not find statistically significant differences in fabric-elasticity relationships between healthy and OI trabecular bone, when only valid ROIs were included and matched for both BV/TV and DA. Therefore, \Add{with the gained knowledge of this study,} HR-pQCT-based hFE analyses used for bone stiffness estimation of healthy individuals can be used for OI diagnosed patients, potentially with a small correction for TMD.
